---
title: "Brazilian Commerce — Exploratory Data Analysis & Data Modelling"
author: ""
date: "2026-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This document covers three sequential tasks:

1.  Loading the raw Olist CSV files and inspecting their structure
2.  Determining primary keys and identifying row-level redundancies for each table
3.  Preparing and exporting cleaned tables to a local PostgreSQL database

------------------------------------------------------------------------

# Libraries

```{r}
library(DataExplorer)
library(tidyverse)
library(purrr)
library(DBI)
library(RPostgres)
```

------------------------------------------------------------------------

# Load Data

Files are loaded by position from the working directory. The expected order is: customers, geolocation, order_items, order_payments, order_reviews, orders, products, sellers, name_trans.

```{r}
file_list <- list.files(pattern = "\\.csv$", full.names = TRUE)

customers      <- read_csv(file_list[1])
geolocation    <- read_csv(file_list[2])
order_items    <- read_csv(file_list[3])
order_payments <- read_csv(file_list[4])
order_reviews  <- read_csv(file_list[5])
orders         <- read_csv(file_list[6])
products       <- read_csv(file_list[7])
sellers        <- read_csv(file_list[8])
name_trans     <- read_csv(file_list[9])
```

------------------------------------------------------------------------

# Database Structure

The two plots below visualise the column-level structure of all tables as a hierarchical network. This serves as a starting point for understanding how tables relate to one another before formal key checking.

```{r fig.height=10, fig.width=25, fig.align='center'}
df_list1 <- list(customers, geolocation, order_items, order_payments)
df_list2 <- list(order_reviews, orders, products, sellers, name_trans)

plot_str(df_list1)
plot_str(df_list2)
```

------------------------------------------------------------------------

# Primary Key Identification

The function below checks whether one or more columns uniquely identify every row in a table. It returns the number of rows, the number of distinct combinations of the specified columns, the number of duplicates, and a boolean indicating whether the columns constitute a valid key.

```{r}
check_key <- function(df, ...) {
  cols <- rlang::ensyms(...)
  col_names <- purrr::map_chr(cols, rlang::as_string)

  n_rows         <- nrow(df)
  n_distinct_rows <- df |> dplyr::distinct(!!!cols) |> nrow()
  n_dupes        <- n_rows - n_distinct_rows

  tibble::tibble(
    columns    = paste(col_names, collapse = " + "),
    n_rows     = n_rows,
    n_distinct = n_distinct_rows,
    n_dupes    = n_dupes,
    is_key     = n_dupes == 0
  )
}
```

## Customers

`customer_id` uniquely identifies each row, but it is not a stable identifier for a person: the same individual receives a new `customer_id` for each order they place. `customer_unique_id` is the persistent identifier across orders.

```{r}
customers |> check_key(customer_id)

n_distinct(customers$customer_id)
n_distinct(customers$customer_unique_id)
```

## Geolocation

No single column or combination of columns uniquely identifies rows. The table contains multiple coordinate entries per zip code prefix, which means it functions as a reference lookup rather than a proper relational table. Latitude and longitude are dropped for the analytical pipeline; only zip code prefix, city, and state are retained as a deduplicated reference.

```{r}
geolocation |> check_key(
  geolocation_zip_code_prefix,
  geolocation_lat,
  geolocation_lng,
  geolocation_city,
  geolocation_state
)

geolocation2 <- geolocation
# There are many Latitude (lat) and longitude (lng) values per zip, city, and state
# Latitude and longitude retained

geolocation <- geolocation |>
  select(geolocation_zip_code_prefix, geolocation_city, geolocation_state) |>
  distinct()
# Latitude and longitude removed
```

## Order Items

The combination of `order_id` and `order_item_id` forms the composite primary key. `order_item_id` indexes individual line items within a single order. `shipping_limit_date` is the deadline by which the seller must hand the item to the carrier.

```{r}
order_items |> check_key(order_id, order_item_id)
```

## Orders

`order_id` is the primary key.

```{r}
orders |> check_key(order_id)
```

## Order Payments

The composite key is `order_id` + `payment_sequential`. `payment_sequential` indexes individual payment entries within a single order, allowing for orders paid using multiple methods (e.g., partial cash and partial credit card produces a `payment_sequential` of 2).

```{r}
order_payments |> check_key(order_id, payment_sequential)
```

## Products

`product_id` is the primary key. Portuguese category names are replaced with English translations via a left join against `name_trans` before export.

```{r}
products |> check_key(product_id)

products <- products |>
  left_join(name_trans, join_by(product_category_name)) |>
  select(-c(product_category_name))
```

## Order Reviews

The composite key is `review_id` + `order_id`. Some orders have more than one associated review. We will keep note of this for subsequent analysis.

```{r}
order_reviews |> check_key(review_id, order_id)
```

## Sellers

`seller_id` is the primary key.

```{r}
sellers |> check_key(seller_id)
```

## Name Translations

`product_category_name` is the primary key.

```{r}
name_trans |> check_key(product_category_name)
```

------------------------------------------------------------------------

# Export to PostgreSQL

Tables with straightforward primary keys (customers, products, sellers) were imported manually via pgAdmin4. The remaining tables are written programmatically via the DBI connector below.

Notes on each table:

-   **orders**: single primary key, no transformation needed
-   **order_reviews**: composite primary key; some orders have multiple reviews
-   **order_items**: composite primary key; a surrogate key will be created in SQL
-   **order_payments**: composite primary key; a surrogate key will be created in SQL
-   **products**: English category names only
-   **geolocation**: deduplicated to zip code prefix, city, and state (\~27,000 rows)
-   **geolocation2**: full table including coordinates, retained for Tableau use

```{r}
connection <- dbConnect(
  Postgres(),
  dbname  = "Brazilian commerce",
  host    = "localhost",
  port    = 5432,
  user    = "postgres",
  password = ""  # Fill in with your own password.
)

tables <- list(
  orders             = orders,
  order_items        = order_items,
  order_payments     = order_payments,
  order_reviews      = order_reviews,
  geolocation        = geolocation,
  geolocation_tableau = geolocation2,
  products           = products,
  dim_customer       = customers,
  dim_seller         = sellers
)

for (name in names(tables)) {
  dbWriteTable(
    conn      = connection,
    name      = name,
    value     = tables[[name]],
    overwrite = TRUE
  )
}
```
