---
title: "Brazilian Commerce — Processing & Cleaning"
author: ""
date: "2026-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(DataExplorer)
library(janitor)
library(naniar)
```

# Load Data

```{r, message=FALSE}
file_list = list.files(path = "../Csvs", pattern = "\\.csv$", full.names = TRUE)

customers      = read_csv(file_list[1], na = c("", "<NA>"))
geolocation    = read_csv(file_list[2])
order_items    = read_csv(file_list[3])
order_payments = read_csv(file_list[4])
order_reviews  = read_csv(file_list[5])
orders         = read_csv(file_list[6])
products       = read_csv(file_list[7])
sellers        = read_csv(file_list[8])
name_trans     = read_csv(file_list[9])

df_list1 = list(customers, geolocation, order_items, order_payments)
df_list2 = list(order_reviews, orders, products, sellers, name_trans)
```

The hierarchical plots below display all tables alongside their column names and types simultaneously, providing a high-level overview of the database structure before we examine each table individually.

```{r}
plot_str(df_list1)
plot_str(df_list2)
```

---

# Processing

### `customers`

```{r, message=FALSE}
glimpse(customers)
# Column types and names look correct.
# Zip code is kept as a string to preserve leading zeros, which are meaningful for identification.

plot_intro(customers)
# No missingness.

customers |> get_dupes()
# No duplicates.
```

Nothing of note; moving on.

---

### `geolocation`

```{r}
glimpse(geolocation)
# Column types and names look correct.

plot_intro(geolocation)
# No missingness.

nrow(geolocation)

geolocation |>
  get_dupes() |>
  distinct()
# ~130,000 duplicate rows detected.

geolocation = geolocation |>
  distinct()
# Duplicates removed.
```

---

### `order_items`

```{r}
glimpse(order_items)
# Column types and names look correct.

plot_intro(order_items)
# No missingness.

order_items |>
  get_dupes() |>
  distinct()
# No duplicates.
```

---

### `order_payments`

```{r}
glimpse(order_payments)
# Column types and names look correct.

plot_intro(order_payments)
# No missingness.

order_payments |>
  get_dupes() |>
  distinct()
# No duplicates.
```

---

### `order_reviews`

```{r}
glimpse(order_reviews)
# Column types and names look correct.

plot_intro(order_reviews)
# Missingness detected — exploring further.

miss_var_summary(order_reviews)
# review_comment_title and review_comment_message have high missingness rates.
# Both columns are retained for potential NLP work; the nature of the missingness
# will be examined in more detail at that stage.

order_reviews |>
  get_dupes() |>
  distinct()
# No duplicates.
```

---

### `orders`

```{r}
glimpse(orders)
# Column types and names look correct.

plot_intro(orders)
# Missingness detected.

miss_var_summary(orders)
# Missingness is confined to date columns only.
```

Date missingness in an e-commerce context is most plausibly caused by logging failures or pipeline issues rather than the values themselves, so MNAR is unlikely. `order_purchase_timestamp` has no missingness — customers successfully placed these orders — but downstream delivery dates are missing, suggesting items were never dispatched or delivered.

```{r}
gg_miss_var(orders, order_status)
```

Cancelled orders partly account for missingness in the delivery date columns, which is expected: cancelled orders are never delivered. Some orders show a shipped status but no delivery date, indicating they were dispatched but had not yet arrived at the time of data collection.

```{r}
orders |>
  mutate(
    purchase_week     = floor_date(order_purchase_timestamp, "week"),
    delivered_missing = is.na(order_delivered_customer_date),
    is_cancelled      = order_status == "canceled"
  ) |>
  group_by(purchase_week) |>
  summarise(
    pct_missing_delivery = mean(delivered_missing) * 100,
    pct_cancelled        = mean(is_cancelled) * 100
  ) |>
  pivot_longer(-purchase_week, names_to = "metric", values_to = "percent") |>
  ggplot(aes(x = purchase_week, y = percent, colour = metric)) +
  geom_line() +
  labs(
    title  = "Delivery date missingness vs cancellation rate over time",
    x      = "Purchase week",
    y      = "%",
    colour = NULL
  )
```

The cancellation rate closely tracks the delivery missingness rate, confirming that cancellations are the primary driver. The spike to 100% missingness at the tail end of the series warrants closer inspection — it is unlikely that every customer cancelled simultaneously, suggesting the data may have become sparse rather than reflecting genuine behaviour.

```{r, warning=FALSE}
weekly_summary = orders |>
  mutate(purchase_week = floor_date(order_purchase_timestamp, "week")) |>
  group_by(purchase_week) |>
  summarise(
    n_orders             = n(),
    pct_missing_delivery = mean(is.na(order_delivered_customer_date)) * 100,
    pct_cancelled        = mean(order_status == "canceled") * 100,
    .groups = "drop"
  )

p1 = weekly_summary |>
  pivot_longer(cols = c(pct_missing_delivery, pct_cancelled),
               names_to = "metric", values_to = "percent") |>
  ggplot(aes(x = purchase_week, y = percent, colour = metric)) +
  geom_line() +
  labs(title = "Rates over time", x = NULL, y = "%", colour = NULL)

p2 = weekly_summary |>
  ggplot(aes(x = purchase_week, y = n_orders)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  labs(title = "Order volume per week", x = "Purchase week", y = "# orders")

library(patchwork)
p1 / p2
```

The volume plot explains the instability: extreme rate swings before 2017-01 and after 2018-07 coincide with periods of very low order counts, where a handful of observations cause large percentage fluctuations.

```{r}
# Confirm where volume collapses at the tail
weekly_summary |>
  filter(purchase_week >= as.Date("2018-07-01")) |>
  select(purchase_week, n_orders, pct_missing_delivery)
```

Volume begins collapsing at the week of 2018-08-26.

```{r}
# Confirm where volume initially stabilises
weekly_summary |>
  filter(purchase_week >= as.Date("2016-08-01") &
         purchase_week <= as.Date("2017-03-01")) |>
  select(purchase_week, n_orders, pct_missing_delivery)
```

Volume stabilises around 2017-01-22. The reliable analysis window is therefore **2017-01-22 to 2018-08-26**.

```{r}
orders |>
  get_dupes() |>
  distinct()
# No duplicates.
```

---

### `products`

```{r, message=FALSE}
glimpse(products)
# Column types and names look correct.

plot_intro(products)
# Missingness detected.

miss_var_summary(products)
# product_category_name, product_name_lenght, product_description_lenght,
# and product_photos_qty all show missingness at n = 610.
# The identical count across all four columns suggests a shared cause:
# the same set of products likely had no listing details filled in by the seller.
```

```{r}
# Verify that the same 610 products are responsible for all missingness
products |>
  group_by(product_id) |>
  summarise(
    miss_cat_name          = sum(is.na(product_category_name)),
    miss_name_len          = sum(is.na(product_name_lenght)),
    miss_description_len   = sum(is.na(product_description_lenght)),
    miss_photos_qty        = sum(is.na(product_photos_qty))
  ) |>
  filter(if_all(starts_with("miss_"), ~ . != 0)) |>
  dim()
# 610 rows confirmed — listing details were not provided for these products.
```

```{r}
products |>
  get_dupes() |>
  distinct()
# No duplicates.
```

---

### `sellers`

```{r}
glimpse(sellers)
# Column types and names look correct.

plot_intro(sellers)
# No missingness.

sellers |> get_dupes()
# No duplicates.
```

---

### `name_trans`

```{r}
glimpse(name_trans)
# Column types and names look correct.

plot_intro(name_trans)
# No missingness.

name_trans |> get_dupes()
# No duplicates.
```
